Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-11-05 manuscript/manuscript.Rmd

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    keep_md: true
    reference_docx: "output_template.docx"
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination. To explore association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting (TODO cite). Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9]

Favipiravir is one of the less explored experimental treatmets, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/IHATMFTC; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of Favipiravir. For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].

High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/ID7F3XIH].
Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the diesease.

An ongoing challenge in evalutating Covid-19 treatments is that the analysis and interpretation of the data is often inappropriate or misleading, most notably interpreting lack of evidence due to small sample size as evidence of no effect [see e.g., @http://zotero.org/users/5567156/items/5PZEEPJV; @http://zotero.org/users/5567156/items/AZDZHLJV].

@@

@@

```{r}
ethics_data <- readxl::read_excel(here::here("manuscript", "ethics.xlsx"), sheet = "Sheet1", na = "NA")
ethics_string <- ethics_data %>% pull(Site) %>% paste0(collapse = ", ")
```

A convenience sample of patients from `r n_hospitals` sites was collected. For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period. The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.   Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy. We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count). Full protocol for data collection is attached in the supplementary material. The study was approved by the Ethical committees of `r ethics_string`.

@@

It shall be noted that the character of the convenience sample did not allow to properly assess 
the association between treatments, markers and patient outcomes,
because the treatments had not been assigned to patients at random (as in a randomized clinical trial), 
but were only observed retrospectively.
Therefore, we could not exclude the possibility of accidental association of treatments to patients' baseline characteristics, 
which could, in turn, demonstrate in spurious association between treatments and outcomes, 
because both of these can be driven by the patients' characteristics. 
Eventhough we attempted to account for possible associations of treatments to patients' baseline characteristics, 
we still could not rule them out, and the results of our analyses should be considered as exploratory.

Since many details of analysis may influence the conclusions made, we performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V] and report results for all the hypothesis tested across multiple different models using both frequentist and Bayesian paradigms. For each model class we worked with several possible sets of adjustments and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC]. All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualisations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].

@@

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighbouring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c). The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction.

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c). AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation. In all models the 'Death' and 'Discharged' states are terminal. In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
```

When markers are used as predictors in both the hidden Markov models and the competing risks model, we impute the missing values using either the maximum of the marker so far or the last-known value of the marker.

Additionally we used joint models for longitudinal and time-to-event data from `stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5]. Those models include a longitudinal component to model the evolution over time of marker for each patient and a time-to-event component which assumes proportional-hazards for an event. The estimated value of the marker is then used as one of the predictors in the time-to-event component.

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features).  We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67]. When there were multiple reasonable ways to evaluate either outcome or a predictor in our dataset, we computed and reported all of those options. As a baseline to compare the scores against we used age and the decade of age as the sole predictor.

@@

data_table_1 <- rbind(
  data$patient_data %>% mutate(group = "All"),
  data$patient_data %>% mutate(group = if_else(ever_hcq, "HCQ", "No HCQ")),
  data$patient_data %>% filter(any_d_dimer) %>% mutate(group = "Measured D-dimer"),
  data$patient_data %>% filter(any_IL_6) %>% mutate(group = "Measured IL-6")
  )

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took Azithromycin` = sum_percent(ever_az),
    `Took Dexamethasone` = sum_percent(ever_dexamethasone),
    `Took Favipiravir` = sum_percent(ever_favipiravir),
    `Took Remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Measured D-dimer","Measured IL-6"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
```

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure &nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization. We also see that the marker levels were not substantially stratified by study site. Those should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.

```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required. Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}

@@

```{r selectedmarkers, fig.cap = "Values of selected markers over the course of the disease. Each line represents a patient, stratified by the worst breathing support required over the hospitalization. Color indicates study sites, notably  The vertical scale is logarithmic. Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
#data$marker_data %>%
# TODO: use SpO2 in AA as SpO2 native (and vice versa)
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

@@

```{r hcq_assoc, message=FALSE, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be asociated with the status of patients upon hospitalization. Comorbidities were asociated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red). HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}

@@

Figure&nbsp;\@ref(fig:hcq_assoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
were less likely to receive Hydroxychloroquine treatment. 
This clearly creates a bias towards a positive effect of Hydroxychloroquine on the outcome, 
which, however, could be false.

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
the score representing the cummulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinin 
was asscociated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

Here, we focus on Hydroxychloroquine and Azithromycin as it is the only treatment with larger sample size. We also investigate Favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented Hydroxychloroquine with Azithromycin while Azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose all but one of the patients receiving Favipiravir also received Hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).

@@

The results of the multiverse analysis for association between Hydrxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - we only report models that we did not find to have immediate problems representing the data well. Results for all models we tested are reported in the supplementary material. Notably, most models report that using Hydroxychloroquine is associated with lower risk of death, we must however bear in mind the potential bias noted in the previous section. Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero, indicating that quite likely better adjustments could drive the effect towards zero. The case of Hydroxychloroquine serves as a "control group" for our other results - since we know from randomized trials that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the observed effects are just a measure of bias in the data. Also, the models we used cannot discern the sign of association between Hydroxychloroquine and in hospital. Additionally, our models either cannot determine the sign of association between Azithromycin and the outcome or even show an increase in risk of death or length of hospitalization.

>>>>

---
title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    keep_md: true
    reference_docx: "output_template.docx"
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination. To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.  Weak evidence to constrain the potential efficacy of Azithromycin and Favipiravir can be extracted from the data. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020 led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteroids, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/NSTX4EZM]. Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/NSTX4EZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

Interestingly, in observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].

Favipiravir is one of the less explored experimental treatments, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA]. Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B]. None of the results allow strong conclusions about benefits or harms of Favipiravir. For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].

High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the disease.

An ongoing challenge in evaluating Covid-19 treatments is that the analysis and interpretation of the data is often inappropriate or misleading, most notably interpreting lack of evidence due to small sample size as evidence of no effect [see e.g., @http://zotero.org/users/5567156/items/5PZEEPJV; @http://zotero.org/users/5567156/items/AZDZHLJV].

@@

The present study aims to describe the outcomes and disease course of hospitalized patients with mild to severe clinical presentation in a multicentric Czech cohort, explore the association between the outcomes and pharmacological interventions and to provide external validation to previously published prognostic models for Covid-19 severity.

@@

```{r}
ethics_data <- readxl::read_excel(here::here("manuscript", "ethics.xlsx"), sheet = "Sheet1", na = "NA")
ethics_string <- ethics_data %>% pull(Site) %>% paste0(collapse = ", ")

n_ecmo <- data$marker_data_wide %>% filter(breathing == "ECMO") %>% pull(patient_id) %>% unique() %>% length()
n_nippv <- data$marker_data_wide %>% filter(breathing == "NIPPV") %>% pull(patient_id) %>% unique() %>% length()
```

A convenience sample of patients from `r n_hospitals` sites was collected. For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period. The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.   Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.

We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count). Full protocol for data collection is attached in the supplementary material. Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we collapsed those categories with mechanical ventilation.

Not all patients developed pneumonia or other symptoms of Covid-19. All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections. Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of attending physician and underlying patient condition. We note that the participating sites were not homogeneous in either patient population or treatment protocols. The choice of treatment was based on the decision of the attending clinician and availability.

The study was approved by the Ethical committees of `r ethics_string`.

@@

The character of the convenience sample does not allow for a proper assessment of
the association between treatments and patient outcomes,
because the treatments had not been assigned to patients at random, 
but were only observed retrospectively. This can be partially remedied by adjusting for patient characteristic in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.

Since many details of analysis may influence the conclusions made, we performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V] and report results for all the hypothesis tested across multiple different models using both frequentist and Bayesian paradigms. For each model class we worked with several possible sets of adjustments. All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualizations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].

```{r}
 # and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC]
```

@@

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c). The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction. All of the hidden Markov models take into account whether best supportive care was initiated and a patient was thus not indicated to progress to more intensive treatment modalities.

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c). AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation. In all models the 'Death' and 'Discharged' states are terminal. In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
```

<!-- When markers are used as predictors in both the hidden Markov models and the competing risks model, we impute the missing values using either the maximum of the marker so far or the last-known value of the marker.   -->

<!-- Additionally we used joint models for longitudinal and time-to-event data via the `stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5]. Those models include a longitudinal component to model the evolution over time of marker for each patient and a time-to-event component which assumes proportional-hazards for an event. The estimated value of the marker is then used as one of the predictors in the time-to-event component.  -->

@@

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features).  We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and its bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67]. When there were multiple reasonable ways to evaluate the outcome or a predictor in our dataset, we computed and reported all of those options. We used two simple scores with age or the decade of age as the sole predictor to have a simple baseline to compare the scores against.

@@

data_table_1 <- rbind(
  data$patient_data %>% mutate(group = "All"),
  data$patient_data %>% mutate(group = if_else(ever_hcq, "HCQ", "No HCQ")),
  data$patient_data %>% filter(ever_favipiravir) %>% mutate(group = "Favipiravir")
  # data$patient_data %>% filter(any_d_dimer) %>% mutate(group = "Measured D-dimer"),
  # data$patient_data %>% filter(any_IL_6) %>% mutate(group = "Measured IL-6")
  )

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took Azithromycin` = sum_percent(ever_az),
    `Took Dexamethasone` = sum_percent(ever_dexamethasone),
    `Took Favipiravir` = sum_percent(ever_favipiravir),
    `Took Remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups. Note that the Favipiravir subgroup is not exclusive with either the HCQ or No HCQ group. IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)

# Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.
```

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization. The limited nature of the data however does not allow for any statistically robust conclusions. We also see that the marker levels were not substantially stratified by study site.  Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured. However we believe this kind of patient-level view is useful to appreciate both the extent of between- and within-patient variability.

```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required. Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}

@@

```{r selectedmarkers, fig.cap = "Values of selected markers over the course of the disease. Each line represents a patient, stratified by the worst breathing support required over the hospitalization. Color indicates study sites. The vertical scale is logarithmic. Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
#data$marker_data %>%
# TODO: use SpO2 in AA as SpO2 native (and vice versa)
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

@@

```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be associated with the status of patients upon hospitalization. Comorbidities were associated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red). HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}

@@

Figure&nbsp;\@ref(fig:hcqassoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
were less likely to receive Hydroxychloroquine treatment. 
This clearly creates a bias towards a positive effect of Hydroxychloroquine on the outcome, 
which, however, could be false.

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinin 
was associated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

Here, we focus on Hydroxychloroquine and Azithromycin as those are the only treatment with larger sample size. We also investigate Favipiravir as it is less well reported in the literature. Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg. Majority of patients complemented Hydroxychloroquine with Azithromycin while Azithromycin was rarely used alone (see Table \@ref(tab:summarytable)). Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.  Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose. All but one of the patients receiving Favipiravir also received Hydroxychloroquine. Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).

@@

The results of the multiverse analysis for association between Hydroxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that we did not find to have immediate problems representing the data well or computational issues. Results for all models we tested are reported in the supplementary material. Notably, most models report that using Hydroxychloroquine is associated with lower risk of death, we must however bear in mind the potential bias noted in the previous section. Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero. This is a weak indication that further adjustments could drive the effect towards zero. We did not attempt to model additional adjustments as the models became computationally intractable. The case of Hydroxychloroquine serves as a "control group" for our other results - since randomized trials give us high confidence that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the associations we observe for Hydroxychloroquine are just a measure of bias in the data. The models we used cannot discern the sign of association between Hydroxychloroquine and length of hospital stay. Additionally, our models either cannot determine the sign of association between Azithromycin and the outcome or even show an increase in risk of death or length of hospitalization. This serves as a weak evidence that a substantial improvement from Azithromycin is unlikely.
